Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks

### 2024-12-12 index.qmd
revamp toy example

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(patchwork)
library(MASS)
library(ggh4x)
library(adtoolbox)
library(broom)
library(knitr)
library(rpart)
library(infotheo)
pm10 <- read_csv(here::here("data", "pm10.csv")) |> 
  filter(!is.na(pm10)) |>    
  mutate(season = as.factor(season))  
```

@@

The rest of the paper is organized as follows: @sec-lit-review reviews the concepts of diagnosing unexpected outcomes and general data quality checks. @sec-plan introduces the concept of analysis validation checks, illustrated with a toy example based on fitness step count data. @sec-method describes the procedure that quantifies how analysis validation checks combined using logical operators can predict unexpected outcomes in an analysis. @sec-pm10-mortality applies this procedure to a larger example that estimates the effect of particulate matter air pollution on daily mortality. @sec-discussion discusses a few key considerations and @sec-conclusion concludes the paper.

@@

Consider a 30-day step count experiment in public health. Subjects are instructed to walk at least 8,000 steps each day, with an expected average of 9,000 steps, tracked by a step counter app. After 30 days, we review the data and examine the number of steps taken each day. With data of this nature, we may expect there to be occasional "low" days due to factors such as forgetting to wear their watch or unfavorable weather conditions limiting outdoor activities.  We may also expect "high" days recorded after an outdoor hike or intense workout. Given the requirements of the study, we expect that for a given subject there would be no more than five days with step counts falling below the 8,000-step threshold ("invalid days"). If there were more than five invalid days, the data from that subject would not be usable for the study.

In this scenario, the outcome is that the number of days with a step count below 8,000 and the expectation is that this outcome would take a value no more than 5. If the expectation is violated, we can consider the potential configurations of the data that could lead to such an anomaly. To diagnose potential reasons why this outcome expectation might fail, we can establish a few analysis validation checks in anticipation of seeing the data. For example, if the average step count is too low, this may suggest that the subject is incapable of taking the required number of steps, potentially leading to the unexpected outcome. Similarly, we can also check the quantile of the step count, if more than a third of the days fall below 8,000, this could indicate an excess of low-count days. Additionally, we may may expect the standard deviation of the step count not to be overly large. These considerations yield the following three analysis validation checks:

-   check 1: the check fails if the mean step count is below 8,200
-   check 2: the check fails if the 30th percentile of the observed step counts is below 8,200
-   check 3: the check fails if the standard deviation of the observed step counts exceeds 2,500.

@@

```{r}
#| label: fig-step-count
#| fig-cap: Number of days with fewer than 8,000 steps across 300 simulated 30-day periods. The orange bars indicate instances where the count exceeds five days, representing an unexpected outcome in this scenario. 
set.seed(1234)
step_count <- tibble(id = 1:300) |> 
  rowwise() |> 
  mutate(
    small_n = rpois(1, 4),
    large_n = rpois(1, 4),
    norm_sd = rlogis(1, 800, scale = 100),
    norm_n = 30 - small_n - large_n,
    step = list(round(c(rnorm(small_n, 4000, 200), 
                        rnorm(large_n, 12000, 200), 
                        rnorm(norm_n, 9000, 300)
                        ), 0)),
    less_than_8000 = sum(step < 8000), 
    unexpect = ifelse(less_than_8000 > 5, 1, 0),
    test1 = ifelse(mean(step) < 8200, 1, 0),
    test2 = ifelse(quantile(step, 0.3, na.rm = TRUE) < 8300, 1, 0),
    test3 = ifelse(sd(step) > 2500, 1, 0),
    ) |>
  ungroup()

# step_count |> 
#   dplyr::select(id, step) |> 
#   unnest(step) |>
#   mutate(day = rep(1:30, 300)) |> 
#   ggplot(aes(x = step, y = id)) + 
#   geom_point(alpha = 0.5) + 
#   theme_bw()

step_count |> 
  ggplot(aes(x = less_than_8000, fill = as.factor(unexpect))) + 
  geom_histogram(binwidth = 0.5) + 
  scale_x_continuous(breaks = seq(0, 13, 1)) +
  scale_fill_brewer(palette = "Dark2") + 
  xlab("Number of days with step count less than 8000") + 
  theme_bw() + 
  theme(legend.position = "none")
```

To simulate this data, three normal distributions are used for the daily step counts: $\mathcal{N}(4000, 200)$ for low days, $\mathcal{N}(12000, 200)$ for high days, and $\mathcal{N}(9000, 300)$ for typical days. The number of low and high days can be simulated from a Poisson distribution with $\lambda = 4$. @fig-step-count displays the number of days with fewer than 8,000 steps across 300 simulated 30-day periods.

@@

To incorporate the effect of multiple checks on the outcome, a logic regression model [@ruczinski_logic_2003] is fitted to the analysis validation checks using the indicator of an unexpected result as the outcome in the model. Originally developed for SNP microarray data, logic regression constructs Boolean combinations of binary variables, in a tree structure, to predict both binary and numerical outcomes. Compared to other tree-based methods for binary-binary prediction, the Boolean combinations from the logic regression model produce a tree structure that can be directly interpreted as the possible combination of checks leading to an unexpected outcome, without the need to invert the tree as required in classic tree-based recursive partitioning methods. The logic regression model is then used to predict the analysis result based on the values of checks, and the prediction is compared to the actual analysis result in order to calculate the precision and recall of the checks.

@@

Returning to the step count example introduced in @sec-toy, the logic regression model is fitted to the three analysis checks described previously to generate the prediction of the unexpected outcome, which is whether the number of days with fewer than 8,000 steps is greater than 5. The predictions from the logic regression model can be compared with the simulated true outcome for calculating the precision and recall metrics. @fig-logic-reg shows the best-fitting logic regression model, which is a combination of check 1 ("mean step count is below 8,200") and check 3 ("standard deviation exceeds 2,500") with an OR operator. In other words, if either check 1 or check 3 is true, then we would predict an unexpected outcome in the analysis (i.e. the number of days with fewer than 8,000 steps is greater than 5).

```{r}
#| label: fig-logic-reg
#| fig.width: 6
#| fig.heigh: 2
#| fig-cap: Logic regression model fitted to the three unit tests (check 1, check 2, check 3). The model suggests using an OR rule to combine check 1 and check 3 to predict the outcome expectation.
fit <- step_count |> fit_logic_reg(unexpect, test1:test3, seed = 1, nleaves = 4)
plot_logicreg(fit)
```

```{r}
fit_regtree <- rpart(unexpect ~ ., data = step_count |> dplyr::select(unexpect:test3), method = "class", 
                     control = rpart.control(cp = 0.01)) 
pred_vec <- predict(fit_regtree) |> as_tibble() |> mutate(.fitted = ifelse(`0` > `1`, 0, 1)) |> dplyr::select(.fitted)
regtree_df <- step_count |> dplyr::select(unexpect) |> bind_cols(pred_vec) |> 
  calc_miscla_rate(unexpect, .fitted) |> 
  mutate(overlapping = multiinformation(step_count[,c(9, 11)])/nrow(step_count), 
         independence = 1 - overlapping) |> 
  calc_metrics(metrics = c("harmonic", "arithmetic"))
```

@tbl-logic-reg presents the calculated precision, recall, and independence for the three individual checks and the combined check rule (check 1 OR check 3) from the logic regression. We also include the metric calculated from fitting a regression tree model to the data to compare the performance of the logic regression model. The harmonic and arithmetic means are included to combine the three measures. The results show that the two checks produced by the logic regression can accurately predict 83.6% cases of all *actual unexpected results* in the simulation data. Furthermore, 82.1% of all *predicted unexpected results* were in fact observed to be unexpected.

```{r}
#| label: tbl-logic-reg
list(tibble(.fitted = step_count$test1, unexpect = step_count$unexpect),
     tibble(.fitted = step_count$test2, unexpect = step_count$unexpect),
     tibble(.fitted = step_count$test3, unexpect = step_count$unexpect)
) |>
  map_dfr(~.x |> calc_miscla_rate(unexpect, .fitted) |> 
            mutate(independence = 1) |> 
            calc_metrics(metrics = c("harmonic", "arithmetic")), .id = "tests") |> 
  mutate(checks = c("check 1", "check 2", "check 3")) |>
  bind_rows(augment(fit) |> 
              calc_miscla_rate(unexpect, .fitted) |> 
              calc_independence() |> 
              calc_metrics(metrics = c("harmonic", "arithmetic")) |> 
              mutate(tests = "check 1 OR check 3") |> 
              dplyr::select(-overlapping)) |> 
  bind_rows(regtree_df |> 
              dplyr::select(-overlapping) |> 
              mutate(tests = "regression tree")) |>
  kable(digits = 3, 
        caption = "Accuracy (precision and recall) and parsimony (independence) metrics for each individual check and for the combined check rule (check 1 OR check 3) derived from the logic regression model. The harmonic and arithmetic means of the three metrics are included to evaluate the quality of the checks in  diagnosing unexpected step counts (more than five days with fewer than 8,000 steps).")
```

For comparison, the regression tree produces a similar prediction to the logic regression, by first splitting on check 1 and then splitting on check 3, and results in the same accuracy and overall score as the logic regression model. However, we argue that the logic regression tree shown in @fig-logic-reg is more interpretable for our purposes because it provides a direct representation of which combinations of analysis checks lead to unexpected outcomes. The logic regression tree is also directly comparable to other diagnostic techniques, which we discuss further in @sec-discussion.

@@

tbl <- tibble(check = c("check 1: sample size <= 200", "check 2: cor(m, PM10) < -0.03",
                        "check 3: cor(m, tmp) > -0.35", "check 4: mortality_outlier", 
                        "(not check 1 AND check 2) AND (check 3 OR check 4)")) |> bind_cols(df |> dplyr::select(-id, -overlapping)) |> rename(`indep.` = independence)

@@

# Conclusion {#sec-conclusion}

TODO

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(patchwork)
library(MASS)
library(ggh4x)
library(adtoolbox)
library(broom)
library(knitr)
library(rpart)
library(infotheo)
library(patchwork)
pm10 <- read_csv(here::here("data", "pm10.csv")) |> 
  filter(!is.na(pm10)) |>    
  mutate(season = as.factor(season))  
```

@@

The rest of the paper is organized as follows: @sec-lit-review reviews the concepts of diagnosing unexpected outcomes and general data quality checks. @sec-plan introduces the concept of analysis validation checks, illustrated with a toy example based on fitness step count data. @sec-method describes the procedure that quantifies how analysis validation checks combined using logical operators can predict unexpected outcomes in an analysis. @sec-pm10-mortality applies this procedure to a larger example that estimates the effect of particulate matter air pollution on daily mortality. @sec-discussion summarises teh paper and discusses a few key considerations.

@@

Consider a 30-day step count experiment in public health. Subjects are instructed to walk at least 8,000 steps each day, with an expected average of 9,000 steps, tracked by a step counter app. After 30 days, we review the data and examine the number of steps taken each day. With data of this nature, we may expect there to be occasional "low" days due to factors such as forgetting to wear their watch or unfavorable weather conditions limiting outdoor activities.  We may also expect "high" days recorded after an outdoor hike or intense workout. Given the requirements of the study, we expect that for a given subject 
the average step count would be between $[8500, 9500]$.

To diagnose potential reasons why this outcome expectation might fail, we can establish a few analysis validation checks in anticipation of seeing the data. For example, we can check the quantile of the step count, if more than a third of the days fall below 8,000, or more than two-thirds exceed 10,000 steps, this could indicate an excess of low-count or high-count days. Similarly, we may may expect the standard deviation of the step count not to be overly large. These considerations yield the following three analysis validation checks that *fails* when:

-   check 1: the 60% quantile of the observed step counts is greater than 10,000
-   check 2: the 40% quantile of the observed step counts is less than 8,000, and
-   check 3: the standard deviation of the observed step counts exceeds 2,500.

@@

```{r}
#| fig-height: 3
#| label: fig-step-count
#| fig-cap: Average step count across 300 simulated 30-day periods. The orange points indicate instances where the average step count fails outside the [8,500, 9,500] interval, representing an unexpected outcome in this scenario. 
set.seed(1234)
step_count <- tibble(id = 1:300) |> 
  rowwise() |> 
  mutate(
    small_n = rpois(1, 8),
    large_n = rpois(1, 8),
    norm_sd = rlogis(1, 800, scale = 100),
    norm_n = 30 - small_n - large_n,
    step = list(round(c(rnorm(small_n, 6000, 200), 
                        rnorm(large_n, 12000, 200), 
                        rnorm(norm_n, 9000, 300)
    ), 0)),
    mean = mean(step, na.rm = TRUE),
    unexpect = ifelse(between(mean, 8500, 9500), 0, 1),
    `q(step, 0.6) > 10000` = ifelse(quantile(step, 0.6, na.rm = TRUE) > 10000, 1, 0),
    `q(step, 0.4) < 8000` = ifelse(quantile(step, 0.4, na.rm = TRUE) < 8000, 1, 0),
    `sd(step) > 2500` = ifelse(sd(step) > 2500, 1, 0),
  ) |>
  ungroup()

step_count |> 
  ggplot(aes(x = mean, y = 1, color = as.factor(unexpect))) + 
  ggbeeswarm::geom_quasirandom() + 
  scale_y_continuous(breaks = seq(0, 13, 1)) +
  scale_color_brewer(palette = "Dark2") + 
  xlab("Average number of steps") + 
  ylab("") + 
  theme_bw() + 
  theme(legend.position = "none", panel.grid.minor = element_blank(), axis.text.y = element_blank())
```

To simulate this data, three normal distributions are used for the daily step counts: $\mathcal{N}(6000, 200)$ for low days, $\mathcal{N}(12000, 200)$ for high days, and $\mathcal{N}(9000, 300)$ for typical days. The number of low and high days can be simulated from a Poisson distribution with $\lambda = 8$. @fig-step-count displays average step count across 300 simulated 30-day periods.

@@

To incorporate the effect of multiple checks on the outcome, a logic regression model [@ruczinski_logic_2003] is fitted to the analysis validation checks. Originally developed for SNP microarray data, logic regression constructs Boolean combinations of binary variables, $\mathcal{L}(X_1, X_2, \cdots, X_n)$, to predict both binary and numerical outcomes. In our case,

* the outcome, $Y$, is a binary variable indicating whether the result is *unexpected* (1) or *expected* (0),
* $X_1, X_2, \cdots, X_n$ represent the binary analysis validation checks, where each check is labeled as 1 if it fails and 0 if it passes.

For classification problems, the logic regression model is reduced to the following form:

$$E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_n)$$
The logical operations of the binary variables $X_1, X_2, \cdots, X_n$, including AND ($\land$), OR ($\lor$), and NOT ($\neg$) and an example of such logic combination can be $X_1 \text{ AND } (X_2 \text{ OR } X_3)$. Simulated annealing is used to explore the search space by minimizing the mis-classification error. The following 6 moves are permitted during the search (as detailed in Figure 2 of @ruczinski_logic_2003): 1) replacing a leaf node, 2) replacing an operator, 3) growing a branch, 4) pruning a branch, 5) splitting a leaf node, and 6) deleting a leaf node.

Compared to other tree-based methods for binary-binary prediction, the Boolean combinations from the logic regression model produce a tree structure that can be directly interpreted as the possible combination of checks leading to an unexpected outcome, without the need to invert the tree as required in classic tree-based recursive partitioning methods. The logic regression model is then used to predict the analysis result based on the values of checks, and the prediction is compared to the actual analysis result in order to calculate the precision and recall of the checks.

@@

Returning to the step count example introduced in @sec-toy, the logic regression model is fitted to the three analysis checks described previously to generate the prediction of the unexpected outcome, whether the average number of step falls within the [8,500, 9,500] interval. The predictions from the logic regression model can be compared with the simulated true outcome for calculating the precision and recall metrics. @fig-logic-reg shows the best-fitting logic regression model as

(quantile(step, 0.6) > 10,000 OR quantile(step, 0.4) < 8,000) AND sd(step) < 2,500

In other words, we would predict an unexpected outcome in the analysis if the standard deviation of the step count is not too large (2,500) and either the 60% quantile of the step count exceeds 10,000 or the 40% quantile of the step count falls below 8,000.

```{r}
#| label: fig-logic-reg
#| fig.width: 8
#| fig.heigh: 2
#| fig-cap: Logic regression model fitted to the three checks. The model suggests the rule (quantile(step, 0.6) > 10,000 OR quantile(step, 0.4) < 8,000) AND sd(step) < 2,500
fit <- step_count |> fit_logic_reg(unexpect, `q(step, 0.6) > 10000`:`sd(step) > 2500`, seed = 1, nleaves = 3)
plot_logicreg(fit)
```

```{r}
fit_regtree <- rpart(unexpect ~ ., data = step_count |> dplyr::select(unexpect:`sd(step) > 2500`), method = "class", 
                     control = rpart.control(cp = 0.01)) 
pred_vec <- predict(fit_regtree) |> as_tibble() |> mutate(.fitted = ifelse(`0` > `1`, 0, 1)) |> dplyr::select(.fitted)
regtree_df <- step_count |> dplyr::select(unexpect) |> bind_cols(pred_vec) |> 
  calc_miscla_rate(unexpect, .fitted) |> 
  mutate(overlapping = multiinformation(step_count[,c(8, 11)])/nrow(step_count), 
         independence = 1 - overlapping) |> 
  calc_metrics(metrics = c("harmonic", "arithmetic"))
```

@tbl-logic-reg presents the calculated precision, recall, and independence for the three individual checks and the check rule found by the logic regression. The harmonic and arithmetic means are included to combine the three measures. The results show that the two checks produced by the logic regression can accurately predict 86.1% cases of all *actual unexpected results* in the simulation data. Furthermore, 43.1% of all *predicted unexpected results* were in fact observed to be unexpected.

```{r}
#| label: tbl-logic-reg
list(tibble(.fitted = step_count$`q(step, 0.6) > 10000`, unexpect = step_count$unexpect),
     tibble(.fitted = step_count$`q(step, 0.4) < 8000`, unexpect = step_count$unexpect),
     tibble(.fitted = step_count$`sd(step) > 2500`, unexpect = step_count$unexpect)
) |>
  map_dfr(~.x |> calc_miscla_rate(unexpect, .fitted) |> 
            mutate(independence = 1) |> 
            calc_metrics(metrics = c("harmonic", "arithmetic"))) |> 
  mutate(checks = c("check 1: q(step, 0.6) > 10000", "check 2: q(step, 0.4) < 8000", "check 3: sd(step) > 2500")) |>
  bind_rows(augment(fit) |> 
              calc_miscla_rate(unexpect, .fitted) |> 
              calc_independence() |> 
              calc_metrics(metrics = c("harmonic", "arithmetic")) |> 
              mutate(checks = "logic regression: (check 1 OR check 2) AND check 3") |> 
              dplyr::select(-overlapping)) |> 
  bind_rows(regtree_df |>
              dplyr::select(-overlapping) |>
              mutate(checks = "regression tree")) |>
  dplyr::select(checks, precision:arithmetic) |> 
  dplyr::rename(`indep.` = independence) |> 
  kable(digits = 3, 
        caption = "Accuracy (precision and recall) and parsimony (independence) metrics for each individual check and the logic regression check rule. The harmonic and arithmetic means of the three metrics are included to evaluate the quality of the checks in  diagnosing unexpected step counts (more than five days with fewer than 8,000 steps).")
```

For comparison, the regression tree produces a similar prediction to the logic regression, however, we argue that the logic regression tree shown in @fig-logic-reg is more interpretable for our purposes because it provides a direct representation of which combinations of analysis checks lead to unexpected outcomes. The logic regression tree is also directly comparable to other diagnostic techniques, which we discuss further in @sec-discussion.

@@

tbl <- tibble(check = c("check 1: sample size <= 200", "check 2: cor(m, PM10) < -0.03",
                        "check 3: cor(m, tmp) > -0.35", "check 4: mortality_outlier", 
                        "logic regression: (not check 1 AND check 2) AND (check 3 OR check 4)")) |> 
  bind_cols(df |> dplyr::select(-id, -overlapping)) |> rename(`indep.` = independence)

@@
